How predictable is extinction?

### 2018-11-28 doc/methods.tex
restructured dir. gsa2018 talk skeleton
readying updates for methods and results

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from Cramer et al. CITATION (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}

@@

%The complement of the hazard function is called the survival function, which describes the probability of a species going extinct after time \(t\). Just as extinction is defined as when the current occurrence is the last occurrence, or \(T = t\), survival is defined as when the current occurrence is less than (in age) than the final occurrence, or \(T > t\). The discrete time survival function is expressed as
%\begin{equation}
%  S(t | X) = P(T > t | X) = \prod{i = 1}{t}(1 - \lambda(i | X).
%  \label{eq:surv}
%\end{equation}

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(0, 10) & \text{if } d = 1 \\
        N(-1, 3) & \text{if } d = 2 \\
        N(0, 3) & \text{if } d > 2 \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             data = data, 
             family = 'binomial',
             prior = normal(c(-1, 0, 0, 0), rep(3, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 3, autoscale = FALSE), 
             chains = 4,
             thin = 4,
             adapt_delta = 0.9999).
\end{verbatim}

@@

One of the issues with our data set is extreme class imbalance; this means that there are many more survival observations than extinction observations. When the classes are not approximately balanced, the base probability of 1 versus 0 is below 0.5. When the intercept term of the logistic regression is sufficiently negative (-2 or less), the only way to obtain a linear predictor value of 0.5 or greater is if the covariate effects are large or there the covariates are at extreme values. The 0.5 default cutpoint for binary prediction is much to high when events are rare. A method for determining the new ``optimal'' cutpoint involves calculating the receiver operating characteristic (ROC) curve from the linear predictor estimates and then identifying the point on the ROC curve which maximizes both true positive rate and the false positive rate. The new cutpoint is then back-calculated from the true positive and false positive rates. Class membership is then predicted given this new cutpoint. This process is done for every posterior draw, as each realization from the posterior predictive distribution has its own optimal cutpoint.

%\section{Variance Partioning and Intraclass Correlation Coeficient}
%
%There are multiple levels and sources of variance in this model. Each of the grouping factors account for some amount of variance in the data. Normalizing that variance by the total variance from all sources, we can see the percent contribution to variance. 
%
%``level 1'' is the ``data level.'' in a regular linear regression problem, this would be the residual variance. that's not calculated in logistic regression, which poses a problem. how do we get this value? simulation?
%
%``level 2'' is the ``time level.'' the logistic regression coefficients \textbf{vary} through time, we are actually estimating that variance in \(\Sigma_{B}\) and \(\tau_{B}\). there is also the non-nested age-based varying-intercept. this source of variance is in \(\Sigma_{a}\) and \(\tau_{a}\).
%
%``level 3'' is the ``phyla level.'' the over-time averages of the regression coefficients further vary between phyla. that variance is part of \(\Sigma_{\alpha}\) and \(\tau_{\alpha}\). there is also the average age effect on extinction probability which is part of \(\Sigma_{\delta}\) and \(\tau_{\delta}\).
%
%that's a lot of variance. 
%
%we want to figure out what the total variance is, so we can determine the percent contribution for each level and segment of variance. the level 1 issue and the nonnestedness pose potential challenges.

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cramer_temp}
  \caption{Comparison of initial temperature estimates from Cramer et al. CITATION (goldenrod) versus the binned values used in this analysis (blue). The initial values are for every 0.1 My while our bins are defined for every 1 My.}
  \label{fig:temp_curve}
\end{figure}

@@

@@

In total, these prior choices can be expressed as
\begin{equation}
  \begin{aligned}
    \mu_{d} &\sim 
      \begin{cases}
        N(-2, 5) & \text{if } d = \text{intercept} \\
        N(-1, 1) & \text{if } d = \text{geo. range} \\
        N(0, 1) & \text{else } \\
      \end{cases}
    \delta &\sim N(0, 1) \\
    \Sigma_{B} &= diag(\tau_{B}) \Omega_{B} diag(\tau_{B}) \\
    \Sigma_{\alpha} &= diag(\tau_{\alpha}) \Omega_{\alpha} diag(\tau_{\alpha}) \\
    \Sigma_{A} &= diag(\tau_{A}) \Omega_{A} diag(\tau_{A}) \\
    \tau_{B} &\sim C^{+}(5) \\
    \tau_{\alpha} &\sim C^{+}(5) \\
    \tau_{A} &\sim C^{+}(5) \\
    \Omega_{B} &\sim LKJ(1) \\
    \Omega_{\alpha} &\sim LKJ(1) \\
    \Omega_{A} &\sim LKJ(1) \\
  \end{aligned}
  \label{eq:priors}
\end{equation}

@@

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             data = data, 
             family = 'binomial',
             prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 3, autoscale = FALSE), 
             chains = 4,
             thin = 4,
             adapt_delta = 0.999999).
\end{verbatim}

@@

### 2018-11-28 doc/results.tex
readying updates for methods and results

Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}). Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points. However, the actual differences between these models are minor as revealed by the wide the standard errors. These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
\begin{table}[ht]
  \centering
  \caption{Model comparison using WAIC and LOOIC}
  \begin{tabular}{ r r r r r }
    \hline
    Model & looic & looic se & waic & waic se \\
    \hline
    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
    \hline
  \end{tabular}
  \label{tab:selection}
\end{table}

As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e. goes extinct). New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.

The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
  \caption{Posterior estimates of the optimal cutpoints for each of the three models. The optimal cutpoint corresponds to the point which maximizes the true and false positive rates; this new cutpoint is used instead of 0.5 when estimating if an observation is the last observation of that species (i.e. goes extinct).}
  \label{fig:cut_point}
\end{figure}

@@

%When our cross-validation estimates are presented for each time interval, 
%% ROC OOS estimate time series
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
%  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
%  \label{fig:fold_auc_time}
%\end{figure}

>>>>

%Our best model, as selected via LOOIC and WAIC, is the parameter rich ``past and vary'' model which includes our historical covariantes and allows all regression parameters to vary through time (Table \ref{tab:selection}). Our best model is an improvement over the next best model by approximately 13 LOOIC and 24 WAIC points. However, the actual differences between these models are minor as revealed by the wide the standard errors. These results mean that while our ``past and vary'' model could be considered the best of the four, all four of the models are very close and only offer incremental improvement but not major differences.
%\begin{table}[ht]
%  \centering
%  \caption{Model comparison using WAIC and LOOIC}
%  \begin{tabular}{ r r r r r }
%    \hline
%    Model & looic & looic se & waic & waic se \\
%    \hline
%    Past and vary & 12784.04 & 178.71 & 12784.04 & 178.71 \\
%    No past but vary & 12819.01 & 178.80 & 12819.01 & 178.80 \\ 
%    Past but no vary & 12847.00 & 179.38 & 12847.00 & 179.38 \\ 
%    No past or vary & 12851.53 & 179.48 & 12851.53 & 179.48 \\ 
%    \hline
%  \end{tabular}
%  \label{tab:selection}
%\end{table}
%
%As described above, our data is extremely unbalanced with respect to the number of extinction events relative to the total number of observations; this reality means that 0.5 is an inappropriate cutpoint for estimating if a species observation is their last (i.e. goes extinct). New, optimal cutpoint were calculated from the ROC curve as the level which maximized the true and false positive rates.
%
%The distributions of optimal cut-off point estimates for each of the four models are broadly similar, with all 50\% credible intervals overlapping.
%
%\begin{figure}[ht]
%  \centering
%  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/cut_plot}
%  \caption{Posterior estimates of the optimal cutpoints for each of the three models. The optimal cutpoint corresponds to the point which maximizes the true and false positive rates; this new cutpoint is used instead of 0.5 when estimating if an observation is the last observation of that species (i.e. goes extinct).}
%  \label{fig:cut_point}
%\end{figure}

@@

When our cross-validation estimates are presented for each time interval,

% ROC OOS estimate time series
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_time}
  \caption{Approximate out-of-sample AUC values calculated for each of the My intervals using five-fold cross-validation of the time series. The AUC of the individual My intervals within each fold is plotted to highlight the heterogentity in performance within and between folds. The AUC estimates from each fold are labeled and are numbered from oldest to youngest.}
  \label{fig:fold_auc_time}
\end{figure}

We can also compare expected out-of-sample performance based on taxonomic group

Finally, expected out-of-sample performance is compared between taxonomic groups over time.
